---
title: "citygrades"
author: "Rajiv Bhatia"
date: "November 1, 2014"
output: 
  rmarkdown::html_vignette:
    fig_caption: no
    fig_height: 6
    fig_width: 6
  
vignette: >
  %\VignetteIndexEntry{City Grades}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---


```{r libraries, echo=FALSE}
library(knitr)
library(tidyr)
library(dplyr)
library(ggvis)
```

```{r data}
library(foodinspections)
data(inspectiondata, package="foodinspections")

```


``` {r organize_data, echo=FALSE}

## Count inspections by business

.icounts <-inspection_tbl %>% group_by(business, type) %>% summarize(count=n()) %>%  data.frame() %>% spread(type,count, fill=0) %>% select(business, routines =Ro, reinspects=Re, complaints =Co) %>% mutate(inspects=routines+reinspects+complaints, retoro=reinspects/routines) %>%  data.frame()

## Count violations by inspections
.vcounts <- violation_tbl %>% group_by(inspection, risklevel) %>% summarize(risk_count=n()) %>%  data.frame() %>% spread(risklevel,risk_count, fill=0)

## Join violation counts to the inspection table
.routines <- inspection_tbl %>% left_join(.vcounts, by="inspection") %>% filter(type=="Ro") %>% select (-type)
.routines[,5:8] [is.na(.routines[,5:8])] <- 0

## Summarize the inspection history for each restaurant
.grades <- .routines %>% filter(score>=0&score<=100) %>% group_by(business) %>% summarize(mean_score=mean(score),high_risk_violations=mean(H)) %>% data.frame() 

## Summarize restaurant history

.restaurants <- business_tbl %>% inner_join (.icounts, by="business") %>% inner_join (.grades, by="business")

```

###Number of Routine Inspections
```{r hist_routines}
.restaurants %>% ggvis(~routines, fill := "blue") %>% layer_histograms()
```

### Average Inspection Score
```{r hist_meanscores}
.restaurants %>% ggvis(~mean_score, fill := "blue") %>% layer_histograms()
```

###Average Number of High Risk Violations
```{r hist_highrisks}
.restaurants %>% ggvis(~high_risk_violations, fill := "blue") %>% layer_histograms()
```

###Ratio of Re-inspections to Routine Inspections

```{r hist_ratio}
.restaurants %>% ggvis(~retoro, fill := "blue") %>% layer_histograms()
```

### Restaurant Scores versus High Risk Violations
```{r plot_scores_violations}
.restaurants %>% ggvis(~high_risk_violations,~mean_score, fill := "blue") %>% layer_points()
```

### Restaurant Scores versus Routine Inspections
```{r plot_scores_routines}
.restaurants %>% ggvis(~routines, ~mean_score, fill := "blue") %>% layer_points()
```

### Restaurant Scores versus Re-inspections
```{r plot_scores_reinspections}
.restaurants %>% ggvis(~reinspects, ~mean_score, fill := "blue") %>% layer_points()
```


### Restaurant Scores versus Complaint Inspections
```{r plot_scores_complaints}
.restaurants %>% ggvis(~complaints, ~mean_score, fill := "blue") %>% layer_points()
```

### Restaurant Scores versus Total Inspections
```{r plot_scores_inspections}
.restaurants %>% ggvis(~inspects, ~mean_score, fill := "blue") %>% layer_points()
```

###Number of Routine Inspections Versus Inspection Scores
```{r plot_scorerange_routineinspections}
.restaurants %>%
  mutate(score_range=cut(mean_score, breaks=c(0,75,80,85,90,95,100))) %>% 
  group_by(score_range) %>% 
  summarize(avg_n_inspections=mean(routines, na.rm=TRUE)) %>%
  ggvis(~score_range,~avg_n_inspections, fill := "blue") %>%
  layer_points()

```

###Number of Re-Inspections Versus Inspection Scores
```{r plot_scorerange_reinspections}

.restaurants %>%
  mutate(score_range=cut(mean_score, breaks=c(0,75,80,85,90,95,100))) %>% 
  group_by(score_range) %>% 
  summarize(avg_n_reinspections=mean(reinspects, na.rm=TRUE)) %>%
  ggvis(~score_range,~avg_n_reinspections, fill := "blue") %>%
  layer_points()

```
