---
title: "Summary Measures of Restaurant Safety"
author: "Rajiv Bhatia"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: no
    fig_height: 6
    fig_width: 6
  
vignette: >
  %\VignetteIndexEntry{Summary Measures of Restaurant Safety}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r libraries, echo=FALSE, message = FALSE}
library(knitr)
library(tidyr)
library(dplyr)
library(corrplot)
library(ggvis)
```


```{r options, echo = FALSE}
opts_chunk$set(echo = FALSE)
options(scipen = 100)
```

### Overview

San Francisco is among a growing number of cities makes restaurant inspection data available to the public. Consumers are now able to access a restaurant's inspection score and history online, both from local governments and from business information services like Yelp.com. While food safety regulations are highly conserved nationally, cities may have diverse procedures for monitoring and scoring restaruants. For example, inspection frequencies may vary from city to city and some cities may allow restaurant to revise a poor inspection score.  As more cities disclose restaurant inspection data, consumers can benefit from uniformly interpretable metrics of restaurant safety peformance. 

This analysis examines three restaurant food safety indicators that could be comparable among cities.

-  Three-year average safety score in the lowest 20th percentile
-  Three-year average ratio of re-inspections to routine inspections greater than one
-  Three-year average number of high risk violations per inspection greater than one from San 

The analysis also examined a cumulative safety score between one and four based on all three indicators. Restaurants can earn a maximum of four points and lose one point for each of the hazard criteria listed above. 

### Data

Food business and inspection data is published weekly by the Department of Public Health. The published data contains records the past three years of safety inspections. The data is accessible at the [Department of Public Health public data FTP site](https://204.68.210.15/). We extracted the following data fields from the public files:

**Business Level Data**

| **business** | Unique identifier for the business. |
| --- | --- |
| **name** | Business name |
| **latitude** | Valid WGS 84 latitude of the business. |
| **longitude** | Valid WGS 84 longitude of the business. |
| **taxcode** | Code representing the category of business for business license purposes |


**Inspection Level Data**

| **business** | Unique identifier for the business. |
| --- | --- |
| **inspection** | Unique identifier for the inspection |
| **date** | Date of inspection |
| **type** | Type of inspection |
| **score** | Restaurant inspection score (for routine inspections only) |


**Violation Level Data**

| **inspection** | Unique identifier for the inspection |
| --- | --- |
| **violation** | Violation type identifier |
| **risklevel** | Health hazard associated with the violation (High, Moderate, Low) |


```{r data}
library(foodinspect)
data(inspectiondata, package="foodinspect")

```


Raw data was used to create a single business level table with summary statistics on inspections and violations for the past three years. Businesses were included if they were tax code classes H24-H27 or H29-H30. This excluded temporary food facilities, mobile facilities, and stadium concessions. A significant number of restaurants had  less than two routine inspections over the three year period.  I excluded these restaurants from the analysis. 

**Restaurant Inspection Table**

| **business** | Unique identifier for the business. |
| --- | --- |
| **name** | Business name |
| **latitude** | Valid WGS 84 latitude of the business. |
| **longitude** | Valid WGS 84 longitude of the business. |
| **taxcode** | Code representing the category of business for business license purposes |
| **routines** | Number of routine inspections |
| **reinspections** | Number of re-inspections |
| **complaints** | Number of complaint inspections |
| **ratio** | Ratio of re-inspections to routine inspections |
| **meanscore** | Average food safety score over three years |
| **highrisks** | Average number of high risk violations per inspection |
| **interval** | Average routine inspection interval |
| **score** | Cumulative safety score |



``` {r organize_data, echo=FALSE}

## Count inspections by business

.icounts <-inspection_tbl %>% 
  group_by(business, type) %>% 
  summarize(count=n()) %>%  
  data.frame() %>% 
  spread(type,count, fill=0) %>% 
  select(business, routines =routine, reinspects=reinspection, complaints =complaint) %>% 
  mutate(inspects=routines+reinspects+complaints, ratio=reinspects/routines) %>%  
  data.frame()

## Count violations by inspections
.vcounts <- violation_tbl %>% 
  group_by(inspection, risklevel) %>% 
  summarize(risk_count=n()) %>%  data.frame() %>% 
  spread(risklevel,risk_count, fill=0)

## Join violation counts to the inspection table
.routines <- inspection_tbl %>% 
  left_join(.vcounts, by="inspection") %>% 
  filter(type=="routine") %>% 
  select (-type)
.routines[,5:8] [is.na(.routines[,5:8])] <- 0

## Summarize the inspection history for each restaurant
.grades <- .routines %>% 
  filter(score>=0&score<=100) %>% 
  group_by(business) %>% 
  summarize(meanscore=mean(score),highrisks=mean(H)) %>% 
  data.frame() 

##Calculate rountine inspection interval

.intervals <- inspection_tbl %>% 
  filter(type=="routine") %>% 
  mutate(date = as.Date(as.character(date),"%Y%m%d")) %>% 
  group_by(business) %>%
  summarize(count =n(), range=(max(date)-min(date)), interval=range/(count-1)) %>%
  filter(count>=2) %>%
  select (business, interval) 

## Summarize restaurant history

.restaurants <- business_tbl %>% 
  inner_join (.icounts, by="business") %>% 
  inner_join (.grades, by="business") %>% 
  inner_join (.intervals, by="business")

# format numerical values

.restaurants[,"ratio"] <- round(.restaurants[,"ratio"], 2)
.restaurants[,"meanscore"] <- round(.restaurants[,"meanscore"], 0)
.restaurants[,"highrisks"] <- round(.restaurants[,"highrisks"], 2)
.restaurants[,"interval"] <- round(.restaurants[,"interval"], 0)

# Create indicator variables and cumulative score

.restaurants[,"c1"] <- as.numeric(.restaurants[,"meanscore"] < quantile(.restaurants[,"meanscore"],probs=0.2))
.restaurants[,"c2"] <- as.numeric(.restaurants[,"highrisks"] > 1)
.restaurants[,"c3"] <- as.numeric(.restaurants[,"ratio"] > 1)
.restaurants[,"score"] <- (4- (.restaurants[,"c1"] + .restaurants[,"c2"] + .restaurants[,"c3"]))
```

---

###Findings

**Number of Routine Inspections a Three Year Period**

```{r hist_routines}
.restaurants %>% ggvis(~routines, fill := "blue") %>% layer_histograms(width=1,center=0)
```

**Time Interval Between Routine Inspections (Days)**

```{r hist_intervals}
.restaurants %>% ggvis(~interval, fill := "blue") %>% layer_histograms(width=90, center=0)
```

**Average Inspection Scores**

```{r hist_meanscores}
.restaurants %>% ggvis(~meanscore, fill := "blue") %>% layer_histograms(width=2)

```

**Average Number of High Risk Violations**

```{r hist_highrisks}
.restaurants %>% ggvis(~highrisks, fill := "blue") %>% layer_histograms(width=0.25)
```

**Ratio of Re-inspections to Routine Inspections**

```{r hist_ratio}
.restaurants %>% ggvis(~ratio, fill := "blue") %>% layer_histograms(width=0.1)
```

**Cumulative Safety Score**


```{r hist_score}
.restaurants %>% ggvis(~score, fill := "blue") %>% layer_histograms(width=0.25)
```

**Restaurants Meeting All Three Cumulative Hazard Criteria**

```{r list_worst, results='asis'}

.worst <-.restaurants %>%
  filter(c1==1&c2==1&c3==1) %>%
  select(name, interval, meanscore, ratio, highrisks) %>%
  arrange(meanscore)
kable(.worst, format="pandoc", row.names=FALSE, align=c("l", "c", "c","c","c"))
           
```
